---
title: "DS6030 Project Part 1: Exploratory Data Analysis"
output: html_document
---

```{r}
library(tidyverse)
library(tidymodels)
library(GGally)
library(ggplot2)
library(knitr)
library(kableExtra)
library(discrim)
library(dplyr) 
library(yardstick)
library(patchwork)
```


```{r}
train_data <- read.csv("train/HaitiPixels.csv")
head(train_data)
```

<span style="font-size: 20px;"><strong>Variable Definitions</strong></span>

<div style="font-size: 14px;">

**Class:** A categorical variable representing the true label for each data point (i.e., what is seen in the aerial image).

• "Vegetation" – likely areas with trees, grass, or plants  
• "Soil" – likely bare earth or dirt  
• "Rooftop" – man-made structures such as buildings  
• "Various Non-Tarp" – miscellaneous surfaces not relevant to identifying displaced persons  
• "Blue Tarp" – the positive class we're trying to detect. These blue tarps likely indicate makeshift shelters created by displaced persons  

**Red, Green, Blue:** Numeric variables ranging from 0 to 255, representing the intensity of each color channel (RGB) in an image.

• 0 represents no intensity (dark) and 255 represents full intensity (pure color)  
• Example: Red = 48, Green = 50, Blue = 230 would appear predominantly blue, possibly indicating a blue tarp  
• Example: Red = 180, Green = 160, Blue = 90 would appear brownish, potentially representing soil or rooftop  

</div>

```{r}
#| fig.width: 9
#| fig.height: 8
# Summarize the data
class_summary <- train_data %>%
  count(Class) %>%
  mutate(
    Percentage = round(100 * n / sum(n), 1),
    FillColor = ifelse(Class == "Blue Tarp", "blue", "#384860")
  )

# Plot with count + percent labels
ggplot(class_summary, aes(x = Class, y = n, fill = FillColor)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(n, " (", Percentage, "%)")), vjust = -0.5) +
  scale_fill_identity() +  # Uses the exact hex colors provided
  theme_minimal() +
  theme(axis.text.x = element_text(size = 12)) + 
  labs(title = "Class Distribution", y = "Count", x = NULL)
```


```{r}
#| fig.width: 12
#| fig.height: 10
# Red
red_bxplt <- ggplot(train_data, aes(x = Class, y = Red)) +
  geom_boxplot(fill = "red") +
  labs(title = "Red Channel Intensity by Class", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1))

# Green
green_bxplt <- ggplot(train_data, aes(x = Class, y = Green)) +
  geom_boxplot(fill = "green") +
  labs(title = "Green Channel Intensity by Class", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1))

# Blue
blue_bxplt <- ggplot(train_data, aes(x = Class, y = Blue)) +
  geom_boxplot(fill = "blue") +
  labs(title = "Blue Channel Intensity by Class", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1))

red_bxplt + green_bxplt + blue_bxplt
```

```{r}
# Create the summary table of medians
median_table <- train_data %>%
  group_by(Class) %>%
  summarise(
    Red = median(Red),
    Green = median(Green),
    Blue = median(Blue)
  ) %>%
  pivot_longer(cols = c(Red, Green, Blue), names_to = "Color", values_to = "Median") %>%
  pivot_wider(names_from = Class, values_from = Median)

# Display table
kable(median_table, caption = "Median RGB Intensities by Class") %>%
  kable_styling(full_width = FALSE, position = "center")
```


```{r}
rgb_dist <- function(data, class_name) {
  class_data <- data %>% filter(Class == class_name)

  # Get max y-count across all three color channels
  max_y <- max(
    hist(class_data$Red, breaks = seq(0, 255, by = 5), plot = FALSE)$counts,
    hist(class_data$Green, breaks = seq(0, 255, by = 5), plot = FALSE)$counts,
    hist(class_data$Blue, breaks = seq(0, 255, by = 5), plot = FALSE)$counts
  )

  g1 <- ggplot(class_data, aes(x = Red)) +
    geom_area(stat = "bin", binwidth = 6, fill = "red", alpha = 1) +
    theme_minimal() +
    labs(x = "Red Channel Intensity", y = "Image Count") +
    coord_cartesian(ylim = c(0, max_y))

  g2 <- ggplot(class_data, aes(x = Green)) +
    geom_area(stat = "bin", binwidth = 6, fill = "green", alpha = 1) +
    theme_minimal() +
    labs(x = "Green Channel Intensity", y = "Image Count") +
    coord_cartesian(ylim = c(0, max_y))

  g3 <- ggplot(class_data, aes(x = Blue)) +
    geom_area(stat = "bin", binwidth = 6, fill = "blue", alpha = 1) +
    theme_minimal() +
    labs(x = "Blue Channel Intensity", y = "Image Count") +
    coord_cartesian(ylim = c(0, max_y))

  (g1 + g2 + g3) +
    plot_annotation(title = paste("Color Breakdown of", class_name, "Images")) &
    theme(plot.title = element_text(size = 20, hjust = 0.5))
}
```

```{r}
#| fig.width: 10
#| fig.height: 6
rgb_dist(train_data, "Blue Tarp")
rgb_dist(train_data, "Vegetation")
rgb_dist(train_data, "Rooftop")
rgb_dist(train_data, "Various Non-Tarp")
rgb_dist(train_data, "Soil")
```

